---
title: "Analyzing your own police department"
author: Andrew Ba Tran, The Washington Post
output: html_document
---

This will walk you through how to analyze a specific police department using data from the FBI Uniform Crime Reporting program and The Washington Post's [Fatal Force](https://www.washingtonpost.com/graphics/investigations/police-shootings-database/) project [[github]](https://github.com/washingtonpost/data-police-shootings). This was part of the [story](): As fatal police shootings increase, more go unreported

## First, identify your police department's ORI code

### 1. Sign up for an API key to access FBI's data

Before you proceed, please [sign up for an API key](https://api.data.gov/signup/) on data.gov.

Here's the [documentation](https://crime-data-explorer.fr.cloud.gov/pages/docApi) for the different FBI crime data endpoints.

### 2. Set up your api key

```{r api_key, eval=F}
fbi_key <- "XXXX"
```

```{r setkey, echo=F}
source("fbi_key.R")
```

### 3. Narrow down your state of choice

```{r state_abb}
state_abb <- "CO"
```

```{r department_data, warning=F, message=F}
# Loading the appropriate libraries

library(tidyverse)
library(lubridate)
library(jsonlite)
library(DT)
library(knitr)

# Importing the json data based off the API
state_df <- fromJSON(paste0("https://api.usa.gov/crime/fbi/sapi/api/agencies/byStateAbbr/", str_to_upper(state_abb), "?API_KEY=", fbi_key))

state_df <- state_df$results

# dropping a few columns for display purposes
state_df %>% 
  select(-state_name, -nibrs, -nibrs_start_date, -region_desc, -division_name) %>% 
  datatable()
```
Above, we have all the agency identification numbers from the state we specified. 

### 4. Export the state dataframe as a spreadsheet

Save the spreadsheet for easier reference, if you'd like.

```{r export1, eval=F}
write_csv(state_df, paste0(state_abb, "_oricodes.csv"), na="")
```

You should be able to narrow down the law enforcement agency of your choice through the `agency_name` column.

Here's how.

### 5. Example: Denver

```{r filter}
# saving the name of an agency we want to look up
department_search <- "denver police"

# filtering out the oricode from the imported state data based on the agency name
ori_only <- state_df %>% 
  filter(str_detect(str_to_lower(agency_name), str_to_lower(department_search))) %>% 
  pull(ori)

ori_only
```

Now we know that `r str_to_title(department_search)`'s ori code is **`r ori_only`**.

## Next, get employment data on the agency

### 1. Import the data from a new API endpoint

You'll need 3 arguments:

* oricode
* beginning year
* end year

```{r employees}
start <- 2015
end <- 2021

emp_url <- paste0("https://api.usa.gov/crime/fbi/sapi/api/police-employment/agencies/", ori_only, "/", start, "/", end, "?API_KEY=", fbi_key)

emp_list <- fromJSON(emp_url)


emp_list <- emp_list$results

# dropping a few columns for display purposes

emp_list_df <- emp_list %>% 
  select(ori, agency_name_edit, year=data_year, population, agency_name_edit, female_officer_ct, male_officer_ct) %>% 
  # calculating total officer count by adding female and male officer counts
  mutate(officer_count=female_officer_ct+male_officer_ct) %>% 
  mutate(year=as.numeric(year))


emp_list_df %>% 
  datatable()
```

Officer counts are necessary to calculate fatal police shootings per officer.

**FBI's agency-level police-involved shootings statistics**

The FBI's API does not give the level of detail necessary to isolate at an agency level officer-involved justifiable homicides from civilian-involved ones. 

Going down this path of analysis will require pulling the full Expanded Homicide [data set](https://crime-data-explorer.fr.cloud.gov/pages/home) from the FBI's [Crime Data Explorer site](https://crime-data-explorer.fr.cloud.gov/pages/home). Alternatively, [Jacob Kaplan](https://jacobdkaplan.weebly.com/)'s Concatenated UCR files from 1976-2020 are also [available](https://www.openicpsr.org/openicpsr/project/100699/version/V11/view). A summarized version from the Post of homicide totals and officer shootings by agency by year can be found [here](data/shr.csv).

## Downloading Fatal Force data specific to an agency

### 1. Import the relationship file between Fatal Force ids and oricodes

Before it recently connected department names to federal ori codes, The Post used internal id numbers to identify law enforcement agencies.

Joining the FBI data with the Post data requires the following crosswalk file.

```{r download1, warning=F, message=F}
fatal_force_ag <- read_csv("https://github.com/washingtonpost/data-police-shootings/raw/master/v2/fatal-police-shootings-agencies.csv")

# showing the first five rows

fatal_force_ag %>% 
  slice(1:5) %>% 
  datatable()
```

This table shows the total shootings in the Post's Fatal Force data by police department starting in 2015 until today.

Grab the `id` for the police agency we're interested in based on the oricode identified above. *Using the oricode as a way to filter is better than the agency name because there may be inconsistent spellings between the Fatal Force and FBI data sets.*

```{r boston_only}
#Using string detect instead of == because there are some agencies we grouped multiple oricodes for simplicity, ie. state police

fatal_force_id <- fatal_force_ag %>% 
  filter(str_detect(ori_only, oricodes)) %>% 
  pull(id) 

fatal_force_id
```

### 1. Download Fatal Force data and filter

Now that we have the department id, we can download the Fatal Force data and filter it to that specific agency.

Some shootings involve more than agency, so we need to transform the data a little to split them up individually.

*Note: `id` from the department data is `agency_ids` in the shootings data*

```{r download2, warning=F, message=F}
fatal_force_df <- read_csv("https://github.com/washingtonpost/data-police-shootings/raw/master/v2/fatal-police-shootings-data.csv")

# Splitting out the incidents with multiple departments so they stand alone and can then be filtered successfully

fatal_force_df <- separate_rows(fatal_force_df, agency_ids, convert = TRUE)

# Now we can filter without worry of leaving any out
fatal_force_df <- fatal_force_df %>% 
  filter(agency_ids==fatal_force_id) 

# Showing a selection of the columns (not all)
fatal_force_df %>% 
  select(id, agency_ids, date, flee_status, armed_with, city, state, name, age, gender, race ) %>% 
  kable()
```

These are all the police shootings tracked by the Post for the specific department.

Now, we count up the shootings by year and join it with the FBI data

### 2. Summarize by year and join with employment data

```{r summarize}
# creating a new year column based on the year
fatal_force_df_annual <- fatal_force_df %>% 
  mutate(year=year(date)) %>%
  count(year, name="fatal_force")

kable(fatal_force_df_annual)

# join with the FBI data

fatal_force_df_joined <- fatal_force_df_annual %>% 
  full_join(emp_list_df) %>% 
  mutate(per1ko=fatal_force/officer_count*1000) %>% 
  filter(year!=2022)

kable(fatal_force_df_joined)

```

### 3. Calculating average annual rate

```{r average}
average_annual <- sum(fatal_force_df_joined$per1ko, na.rm = T)/length(fatal_force_df_joined)

print(paste0(str_to_title(department_search), " average shootings per 1,000 officers: ", round(average_annual, 2)))
```

